\vspace{-1cm} Model linearity breeds contempt: using Bayesian non-linear models to uncover general biogeographical patterns

### 2021-03-18 manuscript/manuscript.tex
Lowering iterations again

Species' distribution models have emerged as one of the most influential methodological advances in ecology and biogeography of the last decades. Useful to understand how populations of species will change along environmental gradients, they have become ecologists' compass to predict the effects of global climate change. That said, uncovering the mechanisms shaping species' realized niches has been one of the main driving forces behind the development of these models. That is, recent efforts have been often focused on understanding which biotic and abiotic factors are good predictors of species' niches---with an increasing effort placed in improving the predictive power of the statistical models. However, we still lack a general understanding of the shape of species' distributions, and much less is known about how these distributions compare to each other across gradients. Here, we use a set of Bayesian non-linear models to uncover the shape of species' realized niches. These models account for all prior knowledge we have regarding their shape, including expert knowledge on species' environmental preferences and physiology. With this approach, we are able to shed light on the true shape of empirical species' distributions. Moreover, they allow us to tackle long-standing hypothesis regarding general biogeographical patterns. In particular, we found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation and their skewness along gradients. Finally, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species. Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.

@@

\subsubsection*{Floristic data}
To complement the aforementioned distribution data, we used a floristic database of most vegetation across Switzerland. This database was build based on expert knowledge and field experience of botanists and ecologists, and contains information regarding species' environmental preferences and physiological traits. Species' environmental preferences in this database can be used to inform distribution models---e.g.~as an informative prior in a Bayesian framework. These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation. These values are provided for a range of 10 climatic variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). On the other hand, the information regarding species' physiological traits represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents. In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species types (e.g.~identifying those species that are neophytes)---and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades). We describe this information in more detail in the Supplementary Table 1.

@@

\subsubsection*{Response curve}
To choose an appropriate response curve, we first need to agree on what we truly know about the system. Given the prior information that we have about the system, we know that species occupy specific geographic ranges; therefore, we know that their distributions have finite variance. While we could also assume that many other factors might influence species' presence in a given site---e.g.~the biotic interactions among specie in the site---we do not necessarily have an \textit{a priori} expectation of how exactly these factors will influence the shape of species' distributions. Therefore, for this baseline model, if all that we are willing to assume about species' realized niches is that these have finite variance, the most conservative assumption and the safest bet---i.e.~the one with the largest entropy---is that they follow a Gaussian distribution. That is, given the presence/absence or abundance $y_{ij}$ of any species $i$ in any given site $j$, and an environmental variable $x_{j}$, we define can species' responses to the environment as
\begin{equation}
\begin{split}
y_{ij} & \sim F\left(p_{ij}\right) \\
\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \gamma_{i} \left(x_{j}-\beta_{i}\right)^2 ,
\end{split}
\label{eq:baseline-response}
\end{equation}
where $F$ is the likelihood function, and $\alpha_i$, $\beta_i^k$, and $\gamma_i$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the an environmental gradient, respectively. Notice that $F$ characterizes a Binomial distribution when considering binary data, and it characterizes an ordered categorical likelihood function when we consider Braun-Blanquet abundance-dominance classes as response variables (see the full description of both models in the Supplementary Methods). For the sake of simplicity, we use only one environmental variable to characterize species' probability distribution. That said, this model can easily be generalized to account for multiple predictors (see Supplementary Methods).

@@

\subsubsection*{Sampling the posterior}
We generated the posterior samples for the Bayesian models with the Hamiltonian Monte Carlo algorithm implementation provided by the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}. Sampling models like the ones described above can be computationally very expensive. This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}). Therefore, we focus on those species for which we have at least 20 occurrences when modelling binary data, which corresponds to the majority of the results of this work. When using ordinal data, we limit our study to a random subset of a 100 species of those with at least 20 occurrences to facilitate the sampling of the models.
% A random subset of a 100 species of those with at least 10 occurrences.

@@

\subsection*{Modifying the baseline model}
We proposed a baseline model that is naive regarding how the data is distributed, and yet accounts for all prior information that we have about the system. Now, we want to modify this model to test the extent to which empirical species' distributions showcase different properties, while preserving both the interpretation of the parameter estimates and the prior information. More specifically, to propose new species' response curves, we follow three key criteria: (i) the probability distribution must have a defined variance and mean, (ii) the Gaussian shape must be a special case of the probability distribution, and (iii) there must be a re-parametrization of the model that allows us to keep the same prior information and interpretable parameters.

\subsubsection*{Heavy-tail response curve}
Distributions with heavy-tails are very common across fields, as they can capture processes such as seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}). Indeed, heavy-tail distributions are pervasive in ecology; for example, species' dispersal patterns have been shown to have heavy-tails due to natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}. Therefore, one might expect these properties to also emerge in alpine communities, where seasonality and dispersal patterns are crucial factors determining species' distributions. To accommodate this feature into our baseline model, we could consider a response curve that follows a generalized error distribution, since the Gaussian shape is a special case of it and contains parameter that regulates the level of kurtosis (Supplementary Figure XX). In particular, we can adapt Eq.~(\ref{eq:baseline-response}) to present this non-linear form as follows:
\begin{equation}
\begin{split}
\text{log}\left(p_{ij}\right) = -\alpha_{i} - \gamma^{\prime}_{i}\, |x_{j}-\beta_{i}|^{\nu_{i}} ,
\end{split}
\label{eq:generalized-response}
\end{equation}
where $\gamma^{\prime}_{i} = g(\gamma_{i}, \nu_{i})$, and $\nu_{i}$ is a parameter that describes the kurtosis of the distribution, which we define as $\nu_{i}\in\left(1, \infty\right)$. Following this, we choose an adaptive prior for this set of new parameter such that $\log\left(\nu_{i}-1\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(2\right)$. Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the corresponding prior distributions (see Supplementary Table 2; Code Availability section; \citealt{nadarajahGeneralizedNormalDistribution2005}). Notice that the Gaussian distribution will naturally emerge when $\nu_i=2$.

Alternatively, we could have used other distributions that present heavy-tails and fulfil the selection criteria described above. For example, the non-standardized Student's t-distributions is an interesting distribution because, as opposed to the generalized error distribution, it allows for heavy tails without generating a cusp at the center (see Supplementary Fig.~XX). However, we avoided using the non-standardized Student's t-distributions because it does not allow for tails that are lighter than normal (e.g.~$\nu_i>2$ in Eq.~\ref{eq:generalized-response}; Supplementary Fig.~XX), and the sampling of the model can be somewhat more challenging (ref).

\subsubsection*{Skewed response curve}
When species experience abiotic or biotic pressures that increase or decrease along an environmental gradient, one might expect their distributions to be skewed in one direction. Likewise, this same skewed shape can also be the result of asymmetric environmental tolerance. One way to accommodate this feature to our models is by considering a skewed normal distribution (Supplementary Figure XX). As for the two cases described above, the Gaussian is a special case of this distribution, and it contains a parameter that controls for the level and direction of `skewness'. Importantly, this distribution presents normal-like tails; therefore, the added skewness does not make additional assumptions regarding how species are distributed along the gradient. To test for the existence of this feature, we modified the species' response curve in Eq.~(\ref{eq:baseline-response}) as
\begin{equation}
\begin{split}
p_{ij} & =\hat{p}_{ij} \left[1+\text{erf}\left(\lambda_{i}\left(x_{j}-\beta^{\prime}_{i}\right)\sqrt{\frac{\gamma^{\prime}_{i}}{2}}\right)\right] \\
\text{log}\left(\hat{p}_{ij}\right) & = -\alpha^{\prime}_{i} - \gamma^{\prime}_{i} \left(x_{j}-\beta^{\prime}_{i}\right)^2 ,
\end{split}
\label{eq:skewed-response}
\end{equation}
where $\alpha^{\prime}_{i} = q_{1}\left(\alpha_{i}, \gamma^{\prime}_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = q_{2}\left(\gamma^{\prime}_{i}, \lambda_{i}\right)$, $\gamma^{\prime}_{i} = q_{3}\left(\gamma_{i}, \lambda_{i}\right)$, and $\lambda_{i}$ is a parameter that describes the skewness of the distribution, which we define as $\lambda_{i}\in\left(-10, 10\right)$. Notice that these boundary conditions for $\lambda_{i}$ are arbitrarily set for computational purposes, as this parameter is theoretically not bounded. The function `erf' is the error function (see \citealt{ashourApproximateSkewNormal2010}). We choose an adaptive prior such that $\lambda_{i}\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$. This model can be re-parametrized following $q_1$, $q_2$ and $q_3$, and set the rest of the prior distributions as described for the baseline model (see Supplementary Table 2; Code Availability section). In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:skewed-response}) when $\lambda_{i}=0$ \citep{ashourApproximateSkewNormal2010}.

One way to test the extend to which species' physiological traits inform the skewness of the distribution of any species $i$ is by characterizing $\lambda_{i}$ as a Gaussian process. That is, we can consider the prior distribution for $\lambda_{i}$ as a multivariate normal with a variance covariance matrix $\Sigma^{\lambda}$. As described before, this variance-covariance matrix can be estimated by the model following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are. Likewise, we could instead use Eq.~(\ref{eq:covariance-complex}) and test the effectiveness of other prior information in explaining the skewness of species' distributions.

\subsubsection*{Heavy-tail and skewed response curve}
Finally, one could also consider a response curve with both kurtosis and skewness. A convenient way to achieve this is by considering a response curve that follows a skewed generalized error distribution (also see XX). This is a combination of the two distributions described above, containing two parameters that control for both the level and direction of kurtosis and skewness. The skewed generalized error distribution can be considered by modifying the species' response curve in Eq.~(\ref{eq:baseline-response}) as
\begin{equation}	
\begin{split}
\text{log}\left(p_{ij}\right) = -\alpha_{i} - \left(\frac{\gamma^{\prime}_{i}\,|x_{j}-\beta^{\prime}_{i}|}{1+\lambda_{i}\,\text{sgn}\left(x_{j}-\beta^{\prime}_{i}\right)}\right)^{\nu_{i}} ,
\end{split}
\label{eq:heavy-skewed-response}
\end{equation}
where $\gamma^{\prime}_{i} = f_{1}\left(\gamma_{i}, \nu_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = f_{2}\left(\gamma_{i}, \beta_{i}, \nu_{i}, \lambda_{i}\right)$, $\nu_{i}$ and $\lambda_i$ are parameters that describe the kurtosis and skewness of the distribution, respectively. $\text{sgn}\left(x\right)$ is the sign function. We define $\nu_i$ and its prior as in Eq.~\ref{eq:generalized-response}. However, $\lambda_i$ is now defined as $\lambda_{i}\in\left(-1, 1\right)$, with an adaptive prior such that $\text{logit}\left(\frac{\lambda_{i}+1}{2}\right)\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$. Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as for the baseline model (see Supplementary Table 2; Code Availability section). Notice that the generalized error distribution (Eq.~\ref{eq:generalized-response}) and the skew normal distribution (Eq.~\ref{eq:skewed-response}) are special cases of Eq.~(\ref{eq:heavy-skewed-response}) when $\lambda_i=0$ and $\nu_i=2$, respectively.

@@

\begin{figure}[h]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=0.9\textwidth]{figures/figure1-prime}
    	  \vspace{0.3cm}
	   \caption{Relationship between mean and variance of species' distributions. Posterior distributions for parameters $\beta_i$ and $\gamma_i$ from Eq.~(\ref{eq:baseline-response}) across species, and the relationship between them. Panel (a) describes the relationship between range size and elevation. Every dot represents the relationship between the mean values for the $\beta_i$ and $\gamma_i$ estimates of the different species. The value in the top-right corner of the plot displays the Pearson's correlation between these parameters calculated across samples of the posterior distributions. Panel (b) describes the $\beta_i$ posterior distribution estimated for all species. Panel (c) describes the $\gamma_i$ posterior distribution estimated for all species. In (b) and (c), the points represent the mean of the posterior distributions, and the corresponding lines characterize the 89\% confidence intervals.}
	   %7.5-5.5
      \label{fig:correlation}
\end{figure}

\begin{figure}[ht]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=0.9\textwidth]{figures/means-Rapopor}
    	  \vspace{0.3cm}
	   \caption{Universality of the relationship between mean and variance of species' distributions. Comparison between how different types of species are mapped in Fig.~\ref{fig:correlation}a. Panel (a) describes the correlation coefficient between $\beta_i$ and $\gamma_i$ for each type of species. Panel (b) shows the differences between neophytes and native species in the way these are distributed along the environmental gradient. Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details). {\color{red} I might move this to the Supplementary Information.}}
	   %7.5-5.5
      \label{fig:universality}
\end{figure}

@@

\begin{figure}[ht]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=0.8\textwidth]{figures/kurto-skew}
    	  \vspace{0.3cm}
	   \caption{Average kurtosis and skewness of species' distributions. Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table XX), the two panels describe the average (a) kurtosis and (b) skewness of distributions, respectively. Panel (a) displays the results obtained by using a response curve that follows a generalized error distribution. Panel (b) displays the results obtained by using a response curve that follows a skewed normal distribution. In both case, the red dotted line indicates the conditions by which species are normally distributed along the environmental axis.}
      \label{fig:kurtosis-skewness}
\end{figure}

Next, we studied the skewness of species distributions using the skewed response curve in Eq.~(\ref{eq:skewed-response}). Based on the estimates for the WAIC values and Akaike weights, this model clearly outperformed the rest (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions. Studying the mean value of the skewness across species (i.e.~parameter $\hat{\lambda}$), we found that species' distributions showed steeper declines towards lower elevations (Fig.~\ref{fig:kurtosis-skewness}). Moreover, species' physiology seems to strongly shape this parameter, which suggest that distribution skewness is an intrinsic property of species (New Fig.~XX)\footref{note1}.

{\color{gray}
[Note: Reports on the hyperparameters describing the different covariance matrices missing at the moment. I think it's interesting, and I will certainly end up adding things here about this. I will also add a third figure showing how lambda varies with the corresponding variance-covariance matrix of species' environmental and physiological traits.]}

>>>>

Species' distribution models have emerged as one of the most influential methodological advances in ecology and biogeography of the last decades. Useful to understand how populations of species will change along environmental gradients, they have become ecologists' compass to predict the effects of global climate change. That said, uncovering the mechanisms shaping species' realized niches has been one of the main driving forces behind the development of these models. That is, recent efforts have been often focused on understanding which biotic and abiotic factors are good predictors of species' niches---with an increasing effort placed in improving the predictive power of the statistical models. However, we still lack a general understanding of the shape of species' distributions, and much less is known about how these distributions compare to each other across gradients. Here, we use a set of Bayesian non-linear models to uncover the shape of species' realized niches. These models account for all prior knowledge we have regarding their shape, including expert knowledge on species' environmental preferences and physiology. With this approach, we are able to uncover the true shape of empirical species' distributions. Moreover, they allow us to tackle long-standing hypothesis regarding general biogeographical patterns. In particular, we found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation and their skewness along gradients. Finally, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species. Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.

@@

\subsubsection*{Floristic data}
To complement the aforementioned distribution data, we used a floristic database of most vegetation across Switzerland. This database was build based on expert knowledge and field experience of botanists and ecologists, and contains information regarding species' environmental preferences and physiological traits. Species' environmental preferences in this database can be used to inform distribution models---e.g.~as an informative prior in a Bayesian framework. These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation. These values are provided for a range of 10 climatic variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). On the other hand, the information regarding species' physiological traits represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents. In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species types (e.g.~identifying those species that are neophytes) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades). We describe this information in more detail in the Supplementary Table 1.

@@

\subsubsection*{Response curve}
To choose an appropriate response curve, we first need to agree on what we truly know about the system. Given the prior information that we have about the system, we know that species occupy specific geographic ranges; therefore, we know that their distributions have finite variance. While we could also assume that many other factors might influence species' presence in a given site---e.g.~the biotic interactions among specie in the site---we do not necessarily have an \textit{a priori} expectation of how exactly these factors will influence the shape of species' distributions. Therefore, for this baseline model, if all that we are willing to assume about species' realized niches is that these have finite variance, the most conservative assumption and the safest bet---i.e.~the one with the largest entropy---is that they follow a Gaussian distribution (Fig~\ref{fig:response}a). That is, given the presence/absence or abundance $y_{ij}$ of any species $i$ in any given site $j$, and an environmental variable $x_{j}$, we can define species' responses to the environment as
\begin{equation}
\begin{split}
y_{ij} & \sim F\left(p_{ij}\right) \\
\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \gamma_{i} \left(x_{j}-\beta_{i}\right)^2 ,
\end{split}
\label{eq:baseline-response}
\end{equation}
where $F$ is the likelihood function, and $\alpha_i$, $\beta_i^k$, and $\gamma_i$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the an environmental gradient, respectively. Notice that $F$ characterizes a Binomial distribution when considering binary data, and it characterizes an ordered categorical likelihood function when we consider Braun-Blanquet abundance-dominance classes as response variables (see the full description of both models in the Supplementary Methods). For the sake of simplicity, we use only one environmental variable to characterize species' probability distribution. That said, this model can easily be generalized to account for multiple predictors (see Supplementary Methods).
\begin{figure}[ht]
  \centering
    \includegraphics[width=1\textwidth]{figures/all-distributions}
	   \caption{Different response curves. Panel (a) shows the probability density function characterized by Eq.~(\ref{eq:baseline-response}) for different values of $\gamma$, when $\mu=0$. Panel (b) shows the probability density function characterized by Eq.~(\ref{eq:generalized-response}) for different values of $\nu$, when $\gamma=1$ and $\mu=0$. Panel (c) shows the probability density function characterized by Eq.~(\ref{eq:skewed-response}) for different values of $\lambda$, when $\gamma=1$ and $\mu=0$. Panel (d) shows the probability density function characterized by Eq.~(\ref{eq:heavy-skewed-response}) for different values of $\lambda$ and $\nu$, when $\gamma=1$ and $\mu=0$. Notice that for each case, we chose $\alpha$ values that normalize the probability distributions.}
      \label{fig:response}
\end{figure}

@@

\subsubsection*{Sampling the posterior}
We generated the posterior samples for the Bayesian models with the Hamiltonian Monte Carlo algorithm implementation provided by the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}. Sampling models like the ones described above can be computationally very intensive. This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}). Therefore, we focus on those species for which we have at least 20 occurrences when modelling binary data. When using ordinal data, we limit our study to a random subset of a 100 species of those with at least 20 occurrences to facilitate the sampling of the models.

@@

\subsection*{Modifying the baseline model}
We proposed a baseline model that is naive regarding how the data is distributed, and yet accounts for all prior information that we have about the system. Now, we want to modify this model to test the extent to which empirical species' distributions showcase different properties, while preserving both the interpretation of the parameter estimates and prior information. To propose new species' response curves, we follow three criteria: (i) the probability distribution must have a defined variance and mean, (ii) the Gaussian shape must be a special case of the probability distribution, and (iii) there must be a re-parametrization of the model that allows us to keep the same prior information and interpretable parameters.

\subsubsection*{Heavy-tail response curve}
Distributions with heavy-tails are very common across fields, as they can capture processes such as seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}). Indeed, heavy-tail distributions are pervasive in ecology; for example, species' dispersal patterns have been shown to have heavy-tails due to natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}. Therefore, one might expect these properties to also emerge in alpine communities, where seasonality and dispersal patterns are crucial factors determining species' distributions. To accommodate this feature into our baseline model, we could consider a response curve that follows a generalized error distribution, since the Gaussian shape is a special case of it and contains parameter that regulates the level of kurtosis (Fig~\ref{fig:response}b). In particular, we can adapt Eq.~(\ref{eq:baseline-response}) to present this non-linear form as follows:
\begin{equation}
\begin{split}
\text{log}\left(p_{ij}\right) = -\alpha_{i} - \gamma^{\prime}_{i}\, |x_{j}-\beta_{i}|^{\nu_{i}} ,
\end{split}
\label{eq:generalized-response}
\end{equation}
where $\gamma^{\prime}_{i} = g(\gamma_{i}, \nu_{i})$, and $\nu_{i}$ is a parameter that describes the kurtosis of the distribution, which we define as $\nu_{i}\in\left(1, \infty\right)$. Following this, we choose an adaptive prior for this set of new parameter such that $\log\left(\nu_{i}-1\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(2\right)$. Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the prior distributions (see Supplementary Table 2; Code Availability section; \citealt{nadarajahGeneralizedNormalDistribution2005}). Notice that the Gaussian distribution will naturally emerge when $\nu_i=2$.

Alternatively, we could have used other distributions that present heavy-tails and fulfil the selection criteria described above. For example, the non-standardized Student's t-distributions is an interesting distribution because, as opposed to the generalized error distribution, it allows for heavy tails without generating a cusp at the center (see Fig~\ref{fig:response}b). However, we avoided using the non-standardized Student's t-distributions because it does not allow for tails that are lighter than normal (e.g.~$\nu_i>2$ in Eq.~\ref{eq:generalized-response}; Fig~\ref{fig:response}b), and the sampling of the model can be somewhat more challenging (ref).

\subsubsection*{Skewed response curve}
When species experience abiotic or biotic pressures that increase or decrease along an environmental gradient, one might expect their distributions to be skewed in one direction. Likewise, this same skewed shape can also emerge as a result of species' asymmetric environmental tolerance. One way to accommodate this feature to our models is by considering a skewed normal distribution (Supplementary Figure XX). As for the two cases described above, the Gaussian is a special case of this distribution, and it contains a parameter that controls for the level and direction of `skewness' (Fig~\ref{fig:response}c). Importantly, this distribution presents normal-like tails; therefore, the added skewness does not make additional assumptions regarding how species are distributed along the gradient. To test for the existence of this feature, we modified Eq.~(\ref{eq:baseline-response}) as
\begin{equation}
\begin{split}
p_{ij} & =\hat{p}_{ij} \left[1+\text{erf}\left(\lambda_{i}\left(x_{j}-\beta^{\prime}_{i}\right)\sqrt{\frac{\gamma^{\prime}_{i}}{2}}\right)\right] \\
\text{log}\left(\hat{p}_{ij}\right) & = -\alpha^{\prime}_{i} - \gamma^{\prime}_{i} \left(x_{j}-\beta^{\prime}_{i}\right)^2 ,
\end{split}
\label{eq:skewed-response}
\end{equation}
where $\alpha^{\prime}_{i} = q_{1}\left(\alpha_{i}, \gamma^{\prime}_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = q_{2}\left(\gamma^{\prime}_{i}, \lambda_{i}\right)$, $\gamma^{\prime}_{i} = q_{3}\left(\gamma_{i}, \lambda_{i}\right)$, and $\lambda_{i}$ is a parameter that describes the skewness of the distribution, which we define as $\lambda_{i}\in\left(-10, 10\right)$. Notice that these boundary conditions for $\lambda_{i}$ are arbitrarily set for computational purposes, as this parameter is theoretically not bounded. The function `erf' is the error function (see \citealt{ashourApproximateSkewNormal2010}). We choose an adaptive prior such that $\lambda_{i}\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$. This model can be re-parametrized following $q_1$, $q_2$ and $q_3$, and set the rest of the prior distributions as described for the baseline model (see Supplementary Table 2; Code Availability section). In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:skewed-response}) when $\lambda_{i}=0$ \citep{ashourApproximateSkewNormal2010}.

One way to test the extend to which species' physiological traits inform the skewness of the distribution of any species $i$ is by characterizing $\lambda_{i}$ as a Gaussian process. That is, we can consider the prior distribution for $\lambda_{i}$ as a multivariate normal with a variance covariance matrix $\Sigma^{\lambda}$. As described before, this variance-covariance matrix can be estimated following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are. Likewise, we could instead use Eq.~(\ref{eq:covariance-complex}) and test the effectiveness of other prior information in explaining the skewness of species' distributions.

\subsubsection*{Heavy-tail and skewed response curve}
Finally, one could consider a response curve with both kurtosis and skewness. A convenient way to achieve this is by using a response curve that follows a skewed generalized error distribution (also see XX). This is a combination of the two distributions described above, containing two parameters that control for both the level and direction of kurtosis and skewness (Fig~\ref{fig:response}d). The skewed generalized error distribution can be considered by modifying the species' response curve in Eq.~(\ref{eq:baseline-response}) as
\begin{equation}	
\begin{split}
\text{log}\left(p_{ij}\right) = -\alpha_{i} - \left(\frac{\gamma^{\prime}_{i}\,|x_{j}-\beta^{\prime}_{i}|}{1+\lambda_{i}\,\text{sgn}\left(x_{j}-\beta^{\prime}_{i}\right)}\right)^{\nu_{i}} ,
\end{split}
\label{eq:heavy-skewed-response}
\end{equation}
where $\gamma^{\prime}_{i} = f_{1}\left(\gamma_{i}, \nu_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = f_{2}\left(\gamma_{i}, \beta_{i}, \nu_{i}, \lambda_{i}\right)$, $\nu_{i}$ and $\lambda_i$ are parameters that describe the kurtosis and skewness of the distribution, respectively. The function $\text{sgn}\left(x\right)$ characterizes the sign function. We define $\nu_i$ and its prior as in Eq.~\ref{eq:generalized-response}. However, $\lambda_i$ is now defined as $\lambda_{i}\in\left(-1, 1\right)$, with an adaptive prior such that $\text{logit}\left(\frac{\lambda_{i}+1}{2}\right)\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$. Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as in the baseline model (see Supplementary Table 2; Code Availability section). Notice that the generalized error distribution (Eq.~\ref{eq:generalized-response}) and the skew normal distribution (Eq.~\ref{eq:skewed-response}) are special cases of Eq.~(\ref{eq:heavy-skewed-response}) when $\lambda_i=0$ and $\nu_i=2$, respectively.

@@

\begin{figure}[h]
  \centering
    \includegraphics[width=0.9\textwidth]{figures/figure1-prime}
    	  \vspace{0.1cm}
	   \caption{Relationship between mean and variance of species' distributions. Posterior distributions for parameters $\beta_i$ and $\gamma_i$ from Eq.~(\ref{eq:baseline-response}) across species, and the relationship between them. Panel (a) describes the relationship between range size and elevation. Every dot represents the relationship between the mean values for the $\beta_i$ and $\gamma_i$ estimates of the different species. The value in the top-right corner of the plot displays the Pearson's correlation between these parameters calculated across samples of the posterior distributions. Panel (b) describes the $\beta_i$ posterior distribution estimated for all species. Panel (c) describes the $\gamma_i$ posterior distribution estimated for all species. In (b) and (c), the points represent the mean of the posterior distributions, and the corresponding lines characterize the 89\% confidence intervals.}
	   %7.5-5.5
      \label{fig:correlation}
\end{figure}

\begin{figure}[ht]
  \centering
    \includegraphics[width=0.9\textwidth]{figures/means-Rapopor}
    	  \vspace{0.1cm}
	   \caption{Universality of the relationship between mean and variance of species' distributions. Comparison between how different types of species are mapped in Fig.~\ref{fig:correlation}a. Panel (a) describes the correlation coefficient between $\beta_i$ and $\gamma_i$ for each type of species. Panel (b) shows the differences between neophytes and native species in the way these are distributed along the environmental gradient. Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details). {\color{red} I might move this to the Supplementary Information.}}
	   %7.5-5.5
      \label{fig:universality}
\end{figure}

@@

\begin{figure}[ht]
  \centering
    \includegraphics[width=0.8\textwidth]{figures/kurto-skew}
    	  \vspace{0.1cm}
	   \caption{Average kurtosis and skewness of species' distributions. Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table XX), the two panels describe the average (a) kurtosis and (b) skewness of distributions, respectively. Panel (a) displays the results obtained by using a response curve that follows a generalized error distribution. Panel (b) displays the results obtained by using a response curve that follows a skewed normal distribution. In both case, the red dotted line indicates the conditions by which species are normally distributed along the environmental axis.}
      \label{fig:kurtosis-skewness}
\end{figure}

Using Eq.~(\ref{eq:skewed-response}), we next studied the skewness of species distributions. Based on the estimates for the WAIC values, this model outperformed the rest (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions. Perhaps most importantly, studying the mean value of the skewness across species (i.e.~parameter $\hat{\lambda}$), we found that species' distributions present steeper declines towards lower elevations (Fig.~\ref{fig:kurtosis-skewness}). Moreover, species' physiology seems to strongly shape this parameter, which suggest that distribution skewness is an intrinsic property of species (New Fig.~XX)\footref{note1}.

Finally, we used a model that allowed for heavy-tails and skewed response curves (Eq.~\ref{eq:heavy-skewed-response}) to understand the relationship between the kurtosis and skewness of distributions. This model outperformed the rest, presenting Akaike weights close to 1 when comparing models (Supplementary Fig.~XX). Comparing the posterior distribution for parameters $\nu_i$ and $\lambda_i$, we found this to be uncorrelated\footref{note1}. However, different types of species seem to present characteristically different distributions based on their kurtosis and skewness. Importantly, this model identified similar relationships between parameters as the baseline model (Supplementary Fig.~XX), and measured similar average levels of kurtosis and skewness than the the other two models (Fig.~\ref{fig:kurtosis-skewness})\footref{note1}.
